Title
========================================================

## Synopsis

## Data Processing
First the raw file is downloaded from the URL, then it is read into memory and converted to a data.table to speedup further operations.

```{r download_and_read_file, cache=TRUE}
library(data.table)
#library(R.utils)
#library(cluster)
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
local.file <- "StormData.csv.bz2"
local.csv <- "StormData.csv"
# download file
if (!file.exists(local.file)) {
  download.file(url, local.file, method="curl", mode="wb")  
}
# read data
df <- read.csv(local.file)
dt <- data.table(df)
```
Next we check the columns of interest.  The column names are listed below. 

* FATALITIES and INJURIES contain the number of people harmed.  
* PROPDMG and PROPDMGEXP represent the significant and exponent of ???? damage.  
* The columns CROPDMG and CROPDMGEXP represent the significant and exponent of ???? damage.
* EVTYPE is the type of event
```{r column_names}
names(dt)
```
To get the correct value of damage, we need to combine the effect of exponent.  The two exponents inlucde integers and shorthand symbols such as K, M, etc.  They also include some invalid symbols.  We convert the exponents to uppercase and create new columns for the damage values.  For exponents of valid symbols (B, M, K) or numeric values, we apply corresponding scaling.  For invalid exponents, no scaling is applied to the damage value.  At last we show the effect of the exponent in two examples.

```{r cleanup_damage}
dt[,table(PROPDMGEXP)]
dt[,table(CROPDMGEXP)]

#define a helper function to convert damage exponent
convert.dmg <- function(significant, exponent, output){
  # change exponent column to upper case
  dt[, exponent := toupper(get(exponent)), with = FALSE]
  # add a new column for output
  dt[, output := get(significant), with = FALSE]
  # apply symbol exponents
  dt[dt[[exponent]] == "B", output := get(significant) * 1e9, with=FALSE]
  dt[dt[[exponent]] == "M", output := get(significant) * 1e6, with=FALSE]  
  dt[dt[[exponent]] == "K", output := get(significant) * 1e3, with=FALSE]  
  # apply integer exponents
  dt[!is.na(as.numeric(dt[[exponent]])), 
     output := get(significant) * 10^as.numeric(get(exponent)), with=FALSE]  
}

convert.dmg("PROPDMG","PROPDMGEXP","prop.dmg")
convert.dmg("CROPDMG","CROPDMGEXP","crop.dmg")

head(dt[PROPDMGEXP == "B", list(PROPDMG, PROPDMGEXP, prop.dmg)])
head(dt[PROPDMGEXP == "5", list(PROPDMG, PROPDMGEXP, prop.dmg)])
```

We also need to cleanup the event types.  As an intial step, we convert it to all upper case and trim white space.  

```{r cleanup_evtype_year}
library(stringr)
dt[, EVTYPE := toupper(str_trim(EVTYPE))]
```

## Results
For intial analysis, we make a summary of persons harmed and total damage per event type.  Then the top 10 events with most serious impact are listed below.
```{r init_analysis}
cols.summary <- c("FATALITIES","INJURIES","prop.dmg","crop.dmg")
summary.event<- dt[,lapply(.SD,sum), by=EVTYPE, 
               .SDcols=cols.summary]
# calculate total persons harmed and total damage
summary.event[,persons.harmed:=FATALITIES+INJURIES]
summary.event[,total.damage:=prop.dmg+crop.dmg]
# list top10
summary.event[order(persons.harmed, decreasing=T)[1:10],
              list(EVTYPE,persons.harmed)]
summary.event[order(total.damage, decreasing=T)[1:10],
              list(EVTYPE, total.damage)]
```

There are only 48 event types defined in the document[1], but the EVTYPE from data file contains several hundreds of types.  We will try to find an official type close to the EVTYPE.

```{r list_evtype}
dt[, length(unique(toupper(EVTYPE)))]

event_names <- unlist(fread("event_name.csv"),use.name=F)
event_names
```

First look the official types.  It is listed in the document [1] and stored in an accompanised file.  A few items cotain two phrases, such as "Hurricane (Typhoon)" or "Frost/Freeze".  We will add more types with splitted phrases.

```

